When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-01-25 doc/paper/vpd_et_paper_PG_v1.tex
done with formatting and spell check, now just ifnal read through

Vapor pressure deficit (VPD) is expected to rise over continents in the future due to the combination of increased temperature and relative humidity drying\explain{Is decreased RH conclusive and universal?- I'm not so sure... is there another citation? I don't think Byrne 2013 shows that.} \citep{Byrne_2013}. Increases in VPD increase the atmospheric demand of evapotranspiration (ET) \citep{Monteith_1965} but also stress plant stomata \citep{Leuning_1990, MEDLYN_2011}.

This leades to two potential but opposing perspectives on how ET adjusts to changes in VPD. The first, a hydrometeorological perspective, is that higher VPD increases atmospheric demand for water from the land surface, and this drives an increase in evapotranspiration (ET). This perspective is particularly relevant because potential evapotranspiration (PET), which is used in many drought indices and hydrometeorological studies, only quantifies atmospheric demand effects. On the other hand, there is another perspective for vegetated surfaces. Plants' stomata have evolved to optimally regulate the exchange of water and carbon, and tend to close in response to increased atmospheric dryness \citep{Ball_1987, Leuning_1990, MEDLYN_2011}.  Therefore, an increase in VPD, particularly in well watered soil conditions, may actually correspond to a decrease in ET because of stomatal closure. This decrease in ET would reduce and potentially cancel (in the case of full stomatal closure) the effects of shifts to atmospheric demand. In other words, the  question ``When does VPD drive or reduce ET?'' can be related to whether plant regulation or atmospheric demand dominates ET.

The ET response to changes in VPD alters water partitioning between the soil and atmosphere. If plant response reduces ET in response to atmospheric drying then soil moisture will be better conserved. This would seem a sensible evolutionary strategy to cope with aridity. If stomata were fully passive \citep [similar to soil pores, e.g. ][]{Or_2013}, increased atmospheric aridity would further reduce soil moisture. In turn, this would further increase aridity as low soil moisture levels increase the Bowen ratio, and in turn increase temperature and reduce atmospheric humidity (gentine et al. GRL 2016, Berg 2016).  This however would not seem to be a sensible strategy for plants from an evolutionary standpoint.

We can use intuition about plant water conservation strategy to hypothesize about ET reponse to changes in VPD. Plants that evolved to conserve water (e.g. arid shrubs) will be more likely to reduce ET with increasing VPD, and plants that have evolved to care little about water (e.g. crops) will be more likely to increase ET with increasing VPD. Atmospheric conditions must matter as well. At the ecosystem scale there are limits to plant water conservation strategies. As atmospheric demand for water (VPD) increases, ecosystems will begin to reach their water conservation limits. At this stage any further increase in VPD will most likely drive an increase in ET, because there is little more the plants can do keep water back.

@@

\section{Data}
\label{data}
We use both meterological and turbulent flux data from the FLUXNET2015 database, including all sites with more than four years of data\explain{Pierre says ``be as precise as possible,'' but not sure what he means}. Each site's plant functional type (PFT) was classified using the International Geosphere-Biosphere Programme vegetation classification scheme \citep{Loveland_1999}. The physical constants used in Section \ref{methods} are only published for five plant functional types (PFTs): crops (CRO), deciduous broadleaf forest (DBF), evergreen needleleaf forest (ENF), grass (GRA), and closed shrub (CSH) (see Table \ref{pft}). There are 56 sites with these plant functional types, and their location is shown in  Figure \ref{map_fig} \explain{map needs to be improved - it's a placeholder for now}.

@@

\section{Methods}
\label{methods}
The Penman-Monteith equation \citep [hereafter PM][]{Monteith_1965} estimates ET as a function of observable atmospheric variables and surface conductances:
\begin{linenomath*}
  \begin{equation}
      ET = \frac{\Delta R + g_a \rho_a c_p VPD}{\Delta + \gamma(1 + \frac{g_a}{g_s})},
  \end{equation}
\end{linenomath*}
where $\Delta$ is the change in saturation vapor pressure with temperature ($\frac{\partial \; e_s}{\partial \; T}$), $R$ is the net radiation minus ground heat flux, $g_a$ is aerodynamic conductance, $\rho_a$ is air density, $c_p$ is specific heat of air at constant pressure, $\gamma$ is the psychometric constant, and $g_s$ is the stomatal conudctance (Table 1). \citet{MEDLYN_2011} developed a model for stomatal conductance ($g_s$) by combining optimal photosynthesis theory \citep{Farquhar_1980, Katul_2010} with empirical approaches, which describes the dependence of $g_s$ to VPD. The result for leaf-scale stomatal conductance was:

@@

where $\Gamma$ is the CO$_2$ compensation point for photosythesis (without dark respiration), and $\lambda$ is the marginal cost of leaf carbon. So, $g_1$ is a leaf-scale term reflecting the willingness of a leaf to trade water for carbon uptake and growth. The higher $g_1$, the more willing a leaf will be willing to release water in the exchange for water.

@@

Note that $uWUE$ is fit on the ecosystem scale in \citet{Zhou_2015} so GPP in \ref{uwue} is really $A\cdot \text{LAI}$. This leads to the cancelation of LAI in addition to uWUE in Equation \ref{medlyn}. Plugging Equation 5 into Equation 1 and rearranging gives a new expression for ET directly as a function of VPD:

@@

With Equation \ref{et}, ET is explicitly a function of environmental variables and two plant-specific constats, the slope parameter ($g_1$), and $uWUE$. The slope parameter is a leaf-scale term related to the willingness of leaves to trade water for new growth. $uWUE$ is semi-empirical ecosystem-scale constant related to how WUE changes with VPD (specifically $VPD^{-1/2}$). Given FLUXNET data described in Section \ref{data}, every term in Equation \ref{et} is known. However, our sampling of sites at the global scale may introduce some deviations of $uWUE$ from those observed in \citet{Zhou_2015}. Also, we wish to include some measure of uncertainty in our analysis to check whether our assumptions and simplifications are reasonable. To account for both mean deviations of $uWUE$ and uncertainty, we will introduce an uncertainty parameter $\sigma$ modifying $uWUE$:

@@

Now, from each FLUXNET observation we can evaluate $\sigma$ at each time step and thus we can evaluate departure from our theory (as a departure from unity). The variability of $\sigma$ across sites and time provides a measure of uncertainty in our model, assumptions, as well as the FLUXNET observations themselves. To correct for differences in sampling between \cite{Zhou_2015} and our data, we set uWUE such that $\overline{\sigma} = 1$. The variability of $\sigma$ then propagates through any uncertainty to our derivative of Equation \ref{et_sigma}:

@@

\subsection{Functional Form of the Sign Term}
\label{sign_func}
First, we explore the variables within the sign term to gain better intuition on the driver of either the increase or reduction of ET with VPD. CO$_2$ concentration ($c_s$) and the pschometric constant ($\gamma$) are relatively constant to a good approximation so that the variability is dominated by $\sigma$ and $VPD$. $uWUE$ could vary with soil moisture but has been shown to be relatively constant except in very dry soil conditions \citep{Zhou_2016, Lin_2015}. We thus further assume that the soil is not too dry so that $uWUE$ can be approximated as a constant (within PFT). This then means that the sign term only depends on VPD for a given PFT and is approximately just a function of $VPD$. We can further determine a critical threshold separating an increase from decrease in ET, such that $VPD_{crit}$ where $\frac{\partial \; ET}{\partial \; VPD} = 0$:

@@

Values of $VPD_{crit}$ as a function of PFT are shown in Table \ref{vpd_crit}. For any values of $VPD$ less than $VPD_{crit}$, ET will decrease with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} < 0$), and for values of $VPD$ greater than $VPD_{crit}$, ET will increase with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} > 0$). In other words, ecosystems can regulate and mitigate evaporative losses up to a VPD limit, above which atmospheric demand is just too high to be entirely compensated by stomatal regulation. We note however that even though ET increases again above the critical threshold, $VPD_{crit}$, ET is still much lower that potential evaporation as stomata are still strongly regulating vapor fluxes. Even in the absence of soil pore evaporation, stomata do not shut down entirely at very high VPD and ET does not go to zero, because stomata are still slightly open to perform some photosynthesis \citep{Ball_1987, Leuning_1990, Medlyn_2011}. In addition, upward xylem transport is necessary to maintain phloem transport and thus carbon allocation (MENCUCCINI) \explain{Not sure citation  - Comparative criteria for models of the vasuclar transport systems of tall trees}.

@@

Differences in the sign term between plant types is exclusively determined by $uWUE$ and the slope parameter ($g_1$). A larger $uWUE$ causes means that at the ecosystem scale $WUE$ decreases more with increasing VPD, so plants lose more water per unit carbon production and are more likely to increase ET with increasing VPD. A tendenacy towards increasing ET reponse with increasing VPD can also be caused by high slope parameters ($g_1$), characteristic of plants that at the leaf scale plants are more willing to trade water for growth. Plants that give water more freely will be much more likely to increase ET with increasing VPD.  Both the aforementioned effects (large $uWUE$, $g_1$) can amplify each other, and generally conspire to shift the sign term towards overall positive values for a given PFT.

This effect of $uWUE$ and $g1$ on the sign term is most apparent by comparing CRO to CSH. CRO has a high slope parameter ($g_1 = 183.1 \, Pa^{1/2}$) and uWUE ($2.32$ -units-) compared to CSH ($g1 = 148.6 \, Pa^{1/2}$, $uWUE=2.11$ -units-). These differences in slope parameter cause opposite ET reponses to changse in VPD between CRO and CSH. ET almost always increases with increasing VPD for CRO, while ET almost always decreases with increasing VPD for CSH (Figure \ref{term3}). CRO, evolved or bred to prioritize GPP, are very willing to trade water for new growth, despite changes in VPD, while CSH are very unwilling to trade water for new growth.

The slope paramter ($g_1$) is additionally a primary determinant of VPD  dependnece for the sign term shown in Figure \ref{term3}. Plants that are less willing to trade water for new growth (small $g_1$) will have strong tendency to reduce ET with increasing VPD, and they will be very effective in reducing ET especially at low VPD.  However, at very high VPD gradients in vapor pressure at the leaf scale will begin to dominate leaf-scale plant strategies (parameterized with $g_1$), so ET reponse will begin to asymtote towards ecosystem-scale values irregardless of leaf-scale strategy (and $g_1$). Therefore, plants with a low $g_1$ will have the largest VPD dependence of ET response because the difference in ET response at low VPD (leaf response dominates) and high VPD (VPD gradient dominates) is largest. This is apparent in the strong VPD dependece of ENF, which has the lowest slope parameter ($g_1=74.31$), in Figure \ref{term3}.

To summarize Figure \ref{term3} and Table \ref{vpd_crit}, CROs are the least water conservative and have the strongest overall tendency to increase ET wirh increasing VPD, while CSH are the most water conservative and have the strongest tendency to decrease ET with increasing VPD.  ENF ($g_1 = 74.31$) has by far the largest VPD dependence of response, while CRO ($g_1 = 183.1$) has the smallest VPD dependence.  ENF are less willing to trade water for new growth at the leaf scale, so the difference between the response at low VPD when leaf response domineates and high VPD when vapor pressure gradients dominate is much greater. CRO on the otherhand are more willing to trade water for new growth, so the leaf scale response is much smaller, and the difference in ET response between low VPD and high VPD is relatively low.

Figure \ref{term3} show clearly that, accoridng to our thoery, for all PFTs except for crops there is frequent occurrence of a negative (plant dominating) ET response to increases in VPD. Therefore, plants are able in most atmospheric conditions to redue ET in response to increased VPD and thus to reduce water loss. To better illustrate this, the ranges of observed environmental VPDs at the FLUXNET sites are plotted parallel to the x-axis. For CSH, VPD is always less than VPD$_{crit}$ so that the plant response dominates and empathizes the water conservative strategy of those plants. For CRO on the otherhand, VPD is almost always higher than VPD$_{crit}$, emphasizing that those plants are water intensive and were actually engineered for photosynthesis rather than water saving. For forests and grass, whether plant response or atmospheric demand dominates is more dependent on the environment, with about half of of observed VPD less than VPD$_{crit}$, i.e. in conditions where plant response dominates. It is also important to note that even when atmospheric demand dominates, the response is still far smaller than it would be for potential evaporation i.e. atmospheric demand only, emphasizing that there is still a strong regulation of evaporative flux by stomata and though the plant xylem. The sign term in this case would just be a constant ($\frac{c_p}{R_{air}} \approx 3.5$), which is far larger than any part of the curves for any PFT. This highlights the deficiencies of PET's ability to capture land response to changes in atmospheric dryness (VPD). Plants are always regulating water exchange from the land surface, even when they reach the limits of they ability to do so. Therefore, the actual land surface reponse to a change in VPD will always me more negative than if the plants are not included in the physics (PET). 
\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig05.pdf}
\caption{Sources of variability for the sign term. Top: sign term as a function of VPD, with $\sigma$ held fixed at 1. The observed range of VPD for each PFT is also shown below the x-axis. Line extent corresponds to 5th and 95th percentiles, while stars denote the location of the 25th, 50th, and 75th percentiles.

@@

\subsection{Functional Form of the Scaling Term}

While the above discussion of the sign of $\frac{\partial \; ET}{\partial \; VPD}$ is important to answer our research question, understaing the overall magnitude of the ET reponse is important to the soil-plant-atmosphere water budgeting. So we now more closely examine the terms that affect how the sign term is scaled:

@@

$\frac{P}{T}$ is an air-density term, which we would expect to vary relatively little compared to aerodynamic conductance and $\Delta$. The psychrometric constant ($\gamma$) is also relatively constant, so the scaling term should be primarily a function of aerodynamic conductance and temperature, through the Clauisu-Clapeyron relationship $\Delta$. This is as expected, given that the aerodynamic conductance represents the efficiency of exchange between the surface and the atmosphere. As aerodynamic conductace  increases, any plant response will be communicated more strongly to the atmosphere (and vice-versa).

$\Delta$'s presence in the scaling term also matches physical intuition. Evaporative cooling will dampen the ability of the atmosphere to take more moisture, because $e_{s}$ decreases with decreasing temperature. The decrease in $e_{s}$ is proportional to $\Delta$ ($\delta e_{s} = \Delta \delta T$). So as $\Delta$ increases, you will get a larger damping of ET due to evaporative cooling.  The functional from of $\Delta$ will be the same across PFT, but the temperature range may vary slightly. In contrast, aerodynamic conductance will vary strongly with PFT due to the importance of surface roughness for aerodynamic conducatance. So most of the differences in scaling between PFT should be in the aerodynamic conductance term.

The control of scaling term variability between PFTS by aerodynamic conductance is confirmed by the data (Figure \ref{scale_vary}). Differences between PFT are almost entiredly due to differences in aerodynamic conductance, rather than differences in observed temperature ranges. The Scaling term for the tree PFTs (DBF, ENF) is generally more than double the scaling terms for the other PFTs which have lower surface roughness and gererally smaller aerodynamic conductance.

Within each PFT, scaling term variability is controlled both by environmental temperature and aerodynamic conductance variability (Figure \ref{scale_vary}). While the observed variability of aerodynamic conductance conrbutes more to scaling term variability than temperature, temperature's contributions are non-negliable. Specifically, the scaling term is generally larger at low temperatures when ET will be weakly damped by Clasius-Clapyeron, and the effect of aerodynamic conductance variability is stronger as it is less weakly damped by Clasius-Clapyeron.

To summarize, variability between PFTs is mostly controlled by systematic differences in aerodynamic conducntace, due to differences in surface roughness, betwen each PFT. In contrast, variability within PFT is also controlled by temperature, through Clasius-Clapyeron. But, aerodynamic conductance variability generally impacts the scaling term more than temperature, even within PFT.

\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig04b.pdf}
\caption{Primary sources of variability for the scaling term, as a function of PFT. Note the change in axis scale for each plot. The 5th-95th percentile range of temperature is plotted at the 5th, 25th, 50th, 75th, and 95th percentiles of aerodynamic conductance, as observed for each PFT's sites.}
\label{scale_vary}
\end{figure}

\subsection{Bulk statistics of ET response to VPD}
\label{stats_sec}
Previous sections only considered idealized scenarios without uncertainty ($\sigma = 1.0$). In this section we will consider bulk statistics, as a function of PFT, of ET reponse when $\sigma$ is allowed to vary and eddy-covariance derived uncertainty is included (Section \ref{methods}).  These bulk statistics of ET reponse confirm our expectations based on the theory presented in Sections \ref{sign_term}-\ref{scale_term} (Table \ref{stats}. For all PFTs, mean ET reponse to increasing VPD is negative. However, ET response evaluated at the average of all variables (e.g. $\sigma$, $T$, $c_s$, $VPD$) is only negative for CSH and GRA. This difference in mean ET reponse as compared to the ET reponse at mean environmental conditions is due to the non-linear nature of the reponse, in which negative reponses are generally much larger magnitude than postive reponses (Figure \ref{sign}).

Regarding the frequency of negative and postive ET response, observations of ENF, GRA and CSH are more frequently characterized by a negative ET reponse to increasing VPD, and are thus more frequently water conservative. DBF and CRO observations have greater frequency of postive ET reponse to increasing VPD, so are less frequently water conservative.

In general the bulk statistics match our expectations based on previous idealzied and theoretical analysis, with the possible exception that inclusion of uncertainty shfits grass towards a slightly more negative reponse than the theory and Figure \ref{sign} would suggest. The bulk statistics motivate a more throrough examination of the structure of uncertainty and more sophsiticated validation of our theory's performance against observations.

>>>>

Vapor pressure deficit (VPD) is expected to rise over continents in the future due to the combination of increased temperature and relative humidity drying\explain{Is decreased RH conclusive and universal?- I'm not so sure... is there another citation? I'm not sure Byrne 2013 shows that.} \citep{Byrne_2013}. Increases in VPD increase the atmospheric demand of evapotranspiration (ET) \citep{Monteith_1965} but also stress plant stomata \citep{Leuning_1990, MEDLYN_2011}.

This leads to two potential but opposing perspectives on how ET adjusts to changes in VPD. The first, a hydrometeorological perspective, is that higher VPD increases atmospheric demand for water from the land surface, and this drives an increase in evapotranspiration (ET). This perspective is particularly relevant because potential evapotranspiration (PET), which is used in many drought indices and hydrometeorological studies, only quantifies atmospheric demand effects. On the other hand, there is another perspective for vegetated surfaces. Plants' stomata have evolved to optimally regulate the exchange of water and carbon, and tend to close in response to increased atmospheric dryness \citep{Ball_1987, Leuning_1990, MEDLYN_2011}.  Therefore, an increase in VPD, particularly in well watered soil conditions, may actually correspond to a decrease in ET because of stomatal closure. This decrease in ET would reduce and potentially cancel (in the case of full stomatal closure) the effects of shifts to atmospheric demand. In other words, the  question ``When does VPD drive or reduce ET?'' can be related to whether plant regulation or atmospheric demand dominates ET.

The ET response to changes in VPD alters water partitioning between the soil and atmosphere. If plant response reduces ET in response to atmospheric drying then soil moisture will be better conserved. This would seem a sensible evolutionary strategy to cope with aridity. If stomata were fully passive \citep [similar to soil pores, e.g. ][]{Or_2013}, increased atmospheric aridity would further reduce soil moisture. In turn, this would further increase aridity as low soil moisture levels increase the Bowen ratio, and in turn increase temperature and reduce atmospheric humidity \citep[gentine et al. GRL 2016][]{Berg_2016}.  This however would not seem to be a sensible strategy for plants from an evolutionary standpoint.

We can use intuition about plant water conservation strategy to hypothesize about ET response to changes in VPD. Plants that evolved to conserve water (e.g. arid shrubs) will be more likely to reduce ET with increasing VPD, and plants that have evolved to care little about water (e.g. crops) will be more likely to increase ET with increasing VPD. Atmospheric conditions must matter as well. At the ecosystem scale there are limits to plant water conservation strategies. As atmospheric demand for water (VPD) increases, ecosystems will begin to reach their water conservation limits. At this stage any further increase in VPD will most likely drive an increase in ET, because there is little more the plants can do keep water back.

@@

\section{Data}
\label{data}
We use both meteorological and turbulent flux data from the FLUXNET2015 database, including all sites with more than four years of data. Each site's plant functional type (PFT) was classified using the International Geosphere-Biosphere Programme vegetation classification scheme \citep{Loveland_1999}. The physical constants used in Section \ref{methods} are only published for five plant functional types (PFTs): crops (CRO), deciduous broadleaf forest (DBF), evergreen needleleaf forest (ENF), grass (GRA), and closed shrub (CSH) (see Table \ref{pft}). There are 56 sites with these plant functional types, and their location is shown in  Figure \ref{map_fig} \explain{map needs to be improved - it's a placeholder for now}.

@@

\section{Methods}
\label{methods}
The Penman-Monteith equation \citep [hereafter PM][]{Monteith_1965} estimates ET as a function of observable atmospheric variables and surface conductances:
\begin{linenomath*}
  \begin{equation}
      ET = \frac{\Delta R + g_a \rho_a c_p VPD}{\Delta + \gamma(1 + \frac{g_a}{g_s})},
  \end{equation}
\end{linenomath*}
where $\Delta$ is the change in saturation vapor pressure with temperature ($\frac{\partial \; e_s}{\partial \; T}$), $R$ is the net radiation minus ground heat flux, $g_a$ is aerodynamic conductance, $\rho_a$ is air density, $c_p$ is specific heat of air at constant pressure, $\gamma$ is the psychometric constant, and $g_s$ is the stomatal conductance (Table 1). \citet{MEDLYN_2011} developed a model for stomatal conductance ($g_s$) by combining optimal photosynthesis theory \citep{Farquhar_1980, Katul_2010} with empirical approaches, which describes the dependence of $g_s$ to VPD. The result for leaf-scale stomatal conductance was:

@@

where $\Gamma$ is the CO$_2$ compensation point for photosynthesis (without dark respiration), and $\lambda$ is the marginal cost of leaf carbon. So, $g_1$ is a leaf-scale term reflecting the willingness of a leaf to trade water for carbon uptake and growth. The higher $g_1$, the more willing a leaf will be willing to open stomata and release water in the exchange for carbon.

@@

Note that $uWUE$ is fit on the ecosystem scale in \citet{Zhou_2015} so GPP in \ref{uwue} is really $A\cdot \text{LAI}$. This leads to the cancellation of LAI in addition to uWUE in Equation \ref{medlyn}. Plugging Equation 5 into Equation 1 and rearranging gives a new expression for ET directly as a function of VPD:

@@

With Equation \ref{et}, ET is explicitly a function of environmental variables and two plant-specific constants, the slope parameter ($g_1$), and $uWUE$. The slope parameter is a leaf-scale term related to the willingness of leaves to trade water for new growth and keep stomata open. $uWUE$ is semi-empirical ecosystem-scale constant related to how WUE changes with VPD (specifically $VPD^{-1/2}$). Given FLUXNET data described in Section \ref{data}, every term in Equation \ref{et} is known. However, our sampling of sites at the global scale may introduce some deviations of $uWUE$ from those observed in \citet{Zhou_2015}. Also, we wish to include some measure of uncertainty in our analysis to check whether our assumptions and simplifications are reasonable. To account for both mean deviations of $uWUE$ and uncertainty, we will introduce an uncertainty parameter $\sigma$ modifying $uWUE$:

@@

Now, from each FLUXNET observation we can evaluate $\sigma$ at each time step:

\begin{linenomath*}
  \begin{equation}
\sigma = - \frac{g_a \gamma c_s \sqrt{VPD} L_v P }{ \left(\text{ ET } ( \Delta + \gamma) - \Delta R - g_a \rho_a c_p VPD\right) 1.6 \; R\; T\; \text{ uWUE } (1 + \frac{g_1}{\sqrt{VPD}})},
    \label{sigma}
  \end{equation}
\end{linenomath*}

and thus we can evaluate departure from our theory (as a departure of $\sigma$ from unity) The variability of $\sigma$ across sites and time provides a measure of uncertainty in our model, assumptions, as well as the FLUXNET observations themselves. To correct for differences in sampling between \cite{Zhou_2015} and our data, we set uWUE such that $\overline{\sigma} = 1$. The variability of $\sigma$ then propagates through any uncertainty to our derivative of Equation \ref{et_sigma}:

@@

\subsection{Functional Form of the Sign Term}
\label{sign_term}
First, we explore the variables within the sign term to gain better intuition on the driver of either the increase or reduction of ET with VPD. CO$_2$ concentration ($c_s$) and the psychometric constant ($\gamma$) are relatively constant to a good approximation so that the variability is dominated by $\sigma$ and $VPD$. $uWUE$ could vary with soil moisture but has been shown to be relatively constant except in very dry soil conditions \citep{Zhou_2016, Lin_2015}. We thus further assume that the soil is not too dry so that $uWUE$ can be approximated as a constant (within PFT). This then means that the sign term only depends on VPD for a given PFT and is approximately just a function of $VPD$. We can further determine a critical threshold separating an increase from decrease in ET, such that $VPD_{crit}$ where $\frac{\partial \; ET}{\partial \; VPD} = 0$:

@@

Values of $VPD_{crit}$ as a function of PFT are shown in Table \ref{vpd_crit}. For any values of $VPD$ less than $VPD_{crit}$, ET will decrease with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} < 0$), and for values of $VPD$ greater than $VPD_{crit}$, ET will increase with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} > 0$). In other words, ecosystems can regulate and mitigate evaporative losses up to a VPD limit, above which atmospheric demand is just too high to be entirely compensated by stomatal regulation. We note however that even though ET increases again above the critical threshold, $VPD_{crit}$, ET is still much lower that potential evaporation as stomata are still strongly regulating vapor fluxes. Even in the absence of soil pore evaporation, stomata do not shut down entirely at very high VPD and ET does not go to zero, because stomata are still slightly open to perform some photosynthesis \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. In addition, upward xylem transport is necessary to maintain phloem transport and thus carbon allocation (MENCUCCINI) \explain{Not sure citation  - Comparative criteria for models of the vasuclar transport systems of tall trees}.

@@

Differences in the sign term between plant types is exclusively determined by $uWUE$ and the slope parameter ($g_1$). A larger $uWUE$ causes means that at the ecosystem scale $WUE$ decreases more with increasing VPD, so plants lose more water per unit carbon production and are more likely to increase ET with increasing VPD. A tendency towards increasing ET response with increasing VPD can also be caused by high slope parameters ($g_1$), characteristic of plants that at the leaf scale plants are more willing to trade water for growth and open stomata. Plants that give water more freely will be much more likely to increase ET with increasing VPD.  Both the aforementioned effects (large $uWUE$, $g_1$) can amplify each other, and generally conspire to shift the sign term towards overall positive values for a given PFT.

This effect of $uWUE$ and $g1$ on the sign term is most apparent by comparing CRO to CSH. CRO has a high slope parameter ($g_1 = 183.1 \, Pa^{1/2}$) and uWUE ($2.32$ -units-) compared to CSH ($g1 = 148.6 \, Pa^{1/2}$, $uWUE=2.11$ -units-). These differences in slope parameter cause opposite ET responses to changes in VPD between CRO and CSH. ET almost always increases with increasing VPD for CRO, while ET almost always decreases with increasing VPD for CSH (Figure \ref{term3}). CRO, evolved or bred to prioritize GPP, are very willing to trade water for new growth, despite changes in VPD, while CSH are very unwilling to trade water for new growth.

The slope parameter ($g_1$) is additionally a primary determinant of VPD  dependence for the sign term shown in Figure \ref{term3}. Plants that are less willing to trade water for new growth (small $g_1$) will have strong tendency to reduce ET with increasing VPD, and they will be very effective in reducing ET especially at low VPD.  However, at very high VPD gradients in vapor pressure at the leaf scale will begin to dominate leaf-scale plant strategies of stomatal closure (parameterized with $g_1$), so ET response will begin to asymptote towards ecosystem-scale values irregardless of leaf-scale strategy and stomatal closure (through $g_1$). Therefore, plants with a low $g_1$ will have the largest VPD dependence of ET response because the difference in ET response at low VPD (leaf response dominates) and high VPD (VPD gradient dominates) is largest. This is apparent in the strong VPD dependence of ENF, which has the lowest slope parameter ($g_1=74.31$), in Figure \ref{term3}.

To summarize Figure \ref{term3} and Table \ref{vpd_crit}, CROs are the least water conservative and have the strongest overall tendency to increase ET with increasing VPD, while CSH are the most water conservative and have the strongest tendency to decrease ET with increasing VPD.  ENF ($g_1 = 74.31$) has by far the largest VPD dependence of response, while CRO ($g_1 = 183.1$) has the smallest VPD dependence.  ENF are less willing to trade water for new growth at the leaf scale, so the difference between the response at low VPD when leaf response dominates and high VPD when vapor pressure gradients dominate is much greater. CRO on the otherhand are more willing to trade water for new growth, so the leaf scale response is much smaller, and the difference in ET response between low VPD and high VPD is relatively low.

Figure \ref{term3} show clearly that, according to our theory, for all PFTs except for crops there is frequent occurrence of a negative (plant dominating) ET response to increases in VPD. Therefore, plants are able in most atmospheric conditions to reduce ET in response to increased VPD and thus to reduce water loss. To better illustrate this, the ranges of observed environmental VPDs at the FLUXNET sites are plotted parallel to the x-axis. For CSH, VPD is always less than VPD$_{crit}$ so that the plant response dominates and empathizes the water conservative strategy of those plants. For CRO on the otherhand, VPD is almost always higher than VPD$_{crit}$, emphasizing that those plants are water intensive and were actually engineered for photosynthesis rather than water saving. For forests and grass, whether plant response or atmospheric demand dominates is more dependent on the environment, with about half of of observed VPD less than VPD$_{crit}$, i.e. in conditions where plant response dominates. It is also important to note that even when atmospheric demand dominates, the response is still far smaller than it would be for potential evaporation i.e. atmospheric demand only, emphasizing that there is still a strong regulation of evaporative flux by stomata and though the plant xylem. The sign term in this case would just be a constant ($\frac{c_p}{R_{air}} \approx 3.5$), which is far larger than any part of the curves for any PFT. This highlights the deficiencies of PET's ability to capture land response to changes in atmospheric dryness (VPD). Plants are always regulating water exchange from the land surface, even when they reach the limits of they ability to do so. Therefore, the actual land surface response to a change in VPD will always me more negative than if the plants are not included in the physics (PET).

\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig05.pdf}
\caption{Sources of variability for the sign term. Top: sign term as a function of VPD, with $\sigma$ held fixed at 1. The observed range of VPD for each PFT is also shown below the x-axis. Line extent corresponds to 5th and 95th percentiles, while stars denote the location of the 25th, 50th, and 75th percentiles.

@@

\subsection{Functional Form of the Scaling Term}
\label{scale_term}
While the above discussion of the sign of $\frac{\partial \; ET}{\partial \; VPD}$ is important to answer our research question, understating the overall magnitude of the ET response is important to the soil-plant-atmosphere water budgeting. So we now more closely examine the terms that affect how the sign term is scaled:

@@

$\frac{P}{T}$ is an air-density term, which we would expect to vary relatively little compared to aerodynamic conductance and $\Delta$. The psychometric constant ($\gamma$) is also relatively constant, so the scaling term should be primarily a function of aerodynamic conductance and temperature, through the Clausius-Clapeyron relationship $\Delta$. This is as expected, given that the aerodynamic conductance represents the efficiency of exchange between the surface and the atmosphere. As aerodynamic conductance  increases, any plant response will be communicated more strongly to the atmosphere (and vice-versa).

$\Delta$'s presence in the scaling term also matches physical intuition. Evaporative cooling will dampen the ability of the atmosphere to take more moisture, because $e_{s}$ decreases with decreasing temperature. The decrease in $e_{s}$ is proportional to $\Delta$ ($\delta e_{s} = \Delta \delta T$). So as $\Delta$ increases, you will get a larger damping of ET due to evaporative cooling.  The functional from of $\Delta$ will be the same across PFT, but the temperature range may vary slightly. In contrast, aerodynamic conductance will vary strongly with PFT due to the importance of surface roughness for aerodynamic conductance. So most of the differences in scaling between PFT should be in the aerodynamic conductance term.

The control of scaling term variability between PFTS by aerodynamic conductance is confirmed by the data (Figure \ref{scale_vary}). Differences between PFT are almost entirely due to differences in aerodynamic conductance, rather than differences in observed temperature ranges. The Scaling term for the tree PFTs (DBF, ENF) is generally more than double the scaling terms for the other PFTs which have lower surface roughness and generally smaller aerodynamic conductance.

Within each PFT, scaling term variability is controlled both by environmental temperature and aerodynamic conductance variability (Figure \ref{scale_vary}). While the observed variability of aerodynamic conductance contributes more to scaling term variability than temperature, temperature's contributions are non-negligible. Specifically, the scaling term is generally larger at low temperatures when ET will be weakly damped by Clausius-Clapeyron, and the effect of aerodynamic conductance variability is stronger as it is less weakly damped by Clausius-Clapeyron.

To summarize, variability between PFTs is mostly controlled by systematic differences in aerodynamic conductance, due to differences in surface roughness, between each PFT. In contrast, variability within PFT is also controlled by temperature, through Clausius-Clapeyron. But, aerodynamic conductance variability generally impacts the scaling term more than temperature, even within PFT.

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{./fig04b.pdf}
\caption{Primary sources of variability for the scaling term, as a function of PFT. Note the change in axis scale for each plot. The 5th-95th percentile range of temperature is plotted at the 5th, 25th, 50th, 75th, and 95th percentiles of aerodynamic conductance, as observed for each PFT's sites.}
\label{scale_vary}
\end{figure}

\subsection{Bulk statistics of ET response to VPD}
\label{stats_sec}
Previous sections only considered idealized scenarios without uncertainty ($\sigma = 1.0$). In this section we will consider bulk statistics, as a function of PFT, of ET response when $\sigma$ is allowed to vary and eddy-covariance derived uncertainty is included (Section \ref{methods}).  These bulk statistics of ET response confirm our expectations based on the theory presented in Sections \ref{sign_term}-\ref{scale_term} (Table \ref{stats}. For all PFTs, mean ET response to increasing VPD is negative. However, ET response evaluated at the average of all variables (e.g. $\sigma$, $T$, $c_s$, $VPD$) is only negative for CSH and GRA. This difference in mean ET response as compared to the ET response at mean environmental conditions is due to the non-linear nature of the response, in which negative responses are generally much larger magnitude than positive responses (Figure \ref{sign}).

Regarding the frequency of negative and positive ET response, observations of ENF, GRA and CSH are more frequently characterized by a negative ET response to increasing VPD, and are thus more frequently water conservative. DBF and CRO observations have greater frequency of positive ET response to increasing VPD, so are less frequently water conservative.

In general the bulk statistics match our expectations based on previous idealized and theoretical analysis, with the possible exception that inclusion of uncertainty shifts grass towards a slightly more negative response than the theory and Figure \ref{sign} would suggest. The bulk statistics motivate a more thorough examination of the structure of uncertainty and more sophisticated validation of our theory's performance against observations.
